---
title: "CodeBook"
author: "Blessing Azubuike"
date: "2024-09-23"
output: pdf_document
---

# CodeBook for UCI HAR Dataset Analysis

## Project Overview
This project analyzes the UCI Human Activity Recognition (HAR) dataset, which contains measurements collected from smartphones to recognize various physical activities performed by subjects.

## Variables Description

### Data Structure
The final dataset contains the following variables:

| Variable Name                                           | Description                                           | Units            |
|---------------------------------------------------------|-------------------------------------------------------|------------------|
| `subject`                                               | Unique identifier for each subject                    | N/A              |
| `activity_code`                                        | Numerical code representing the activity              | N/A              |
| `activity_name`                                        | Descriptive name of the activity                      | N/A              |
| `TimeBodyAcc-Mean-X`                                  | Mean body acceleration along the X-axis               | g (gravity)      |
| `TimeBodyAcc-Mean-Y`                                  | Mean body acceleration along the Y-axis               | g (gravity)      |
| `TimeBodyAcc-Mean-Z`                                  | Mean body acceleration along the Z-axis               | g (gravity)      |
| `TimeBodyAcc-StandardDeviation-X`                     | Standard deviation of body acceleration along the X-axis | g (gravity)   |
| `TimeBodyAcc-StandardDeviation-Y`                     | Standard deviation of body acceleration along the Y-axis | g (gravity)   |
| `TimeBodyAcc-StandardDeviation-Z`                     | Standard deviation of body acceleration along the Z-axis | g (gravity)   |
| `TimeGravityAcc-Mean-X`                               | Mean gravity acceleration along the X-axis            | g (gravity)      |
| `TimeGravityAcc-Mean-Y`                               | Mean gravity acceleration along the Y-axis            | g (gravity)      |
| `TimeGravityAcc-Mean-Z`                               | Mean gravity acceleration along the Z-axis            | g (gravity)      |
| `TimeGravityAcc-StandardDeviation-X`                  | Standard deviation of gravity acceleration along the X-axis | g (gravity)   |
| `TimeGravityAcc-StandardDeviation-Y`                  | Standard deviation of gravity acceleration along the Y-axis | g (gravity)   |
| `TimeGravityAcc-StandardDeviation-Z`                  | Standard deviation of gravity acceleration along the Z-axis | g (gravity)   |
| `TimeBodyAccJerk-Mean-X`                              | Mean of body acceleration jerk along the X-axis       | g (gravity)      |
| `TimeBodyAccJerk-Mean-Y`                              | Mean of body acceleration jerk along the Y-axis       | g (gravity)      |
| `TimeBodyAccJerk-Mean-Z`                              | Mean of body acceleration jerk along the Z-axis       | g (gravity)      |
| `TimeBodyAccJerk-StandardDeviation-X`                 | Standard deviation of body acceleration jerk along the X-axis | g (gravity)   |
| `TimeBodyAccJerk-StandardDeviation-Y`                 | Standard deviation of body acceleration jerk along the Y-axis | g (gravity)   |
| `TimeBodyAccJerk-StandardDeviation-Z`                 | Standard deviation of body acceleration jerk along the Z-axis | g (gravity)   |
| `TimeBodyGyro-Mean-X`                                 | Mean body gyroscope measurement along the X-axis      | rad/s            |
| `TimeBodyGyro-Mean-Y`                                 | Mean body gyroscope measurement along the Y-axis      | rad/s            |
| `TimeBodyGyro-Mean-Z`                                 | Mean body gyroscope measurement along the Z-axis      | rad/s            |
| `TimeBodyGyro-StandardDeviation-X`                    | Standard deviation of body gyroscope measurement along the X-axis | rad/s         |
| `TimeBodyGyro-StandardDeviation-Y`                    | Standard deviation of body gyroscope measurement along the Y-axis | rad/s         |
| `TimeBodyGyro-StandardDeviation-Z`                    | Standard deviation of body gyroscope measurement along the Z-axis | rad/s         |
| `TimeBodyGyroJerk-Mean-X`                             | Mean of gyroscope jerk measurement along the X-axis    | rad/s            |
| `TimeBodyGyroJerk-Mean-Y`                             | Mean of gyroscope jerk measurement along the Y-axis    | rad/s            |
| `TimeBodyGyroJerk-Mean-Z`                             | Mean of gyroscope jerk measurement along the Z-axis    | rad/s            |
| `TimeBodyGyroJerk-StandardDeviation-X`                | Standard deviation of gyroscope jerk measurement along the X-axis | rad/s         |
| `TimeBodyGyroJerk-StandardDeviation-Y`                | Standard deviation of gyroscope jerk measurement along the Y-axis | rad/s         |
| `TimeBodyGyroJerk-StandardDeviation-Z`                | Standard deviation of gyroscope jerk measurement along the Z-axis | rad/s         |
| `TimeBodyAccMag-Mean`                                 | Mean magnitude of body acceleration                    | g (gravity)      |
| `TimeBodyAccMag-StandardDeviation`                     | Standard deviation of magnitude of body acceleration   | g (gravity)      |
| `TimeGravityAccMag-Mean`                              | Mean magnitude of gravity acceleration                 | g (gravity)      |
| `TimeGravityAccMag-StandardDeviation`                  | Standard deviation of magnitude of gravity acceleration | g (gravity)     |
| `TimeBodyAccJerkMag-Mean`                             | Mean magnitude of body acceleration jerk               | g (gravity)      |
| `TimeBodyAccJerkMag-StandardDeviation`                 | Standard deviation of magnitude of body acceleration jerk | g (gravity)   |
| `TimeBodyGyroMag-Mean`                                | Mean magnitude of body gyroscope measurement           | rad/s            |
| `TimeBodyGyroMag-StandardDeviation`                    | Standard deviation of magnitude of body gyroscope measurement | rad/s        |
| `TimeBodyGyroJerkMag-Mean`                            | Mean magnitude of gyroscope jerk measurement           | rad/s            |
| `TimeBodyGyroJerkMag-StandardDeviation`                | Standard deviation of magnitude of gyroscope jerk measurement | rad/s        |
| `FrequencyBodyAcc-Mean-X`                              | Mean body acceleration in frequency domain along the X-axis | g (gravity)   |
| `FrequencyBodyAcc-Mean-Y`                              | Mean body acceleration in frequency domain along the Y-axis | g (gravity)   |
| `FrequencyBodyAcc-Mean-Z`                              | Mean body acceleration in frequency domain along the Z-axis | g (gravity)   |
| `FrequencyBodyAcc-StandardDeviation-X`                 | Standard deviation of body acceleration in frequency domain along the X-axis | g (gravity) |
| `FrequencyBodyAcc-StandardDeviation-Y`                 | Standard deviation of body acceleration in frequency domain along the Y-axis | g (gravity) |
| `FrequencyBodyAcc-StandardDeviation-Z`                 | Standard deviation of body acceleration in frequency domain along the Z-axis | g (gravity) |
| `FrequencyBodyAccJerk-Mean-X`                          | Mean of body acceleration jerk in frequency domain along the X-axis | g (gravity) |
| `FrequencyBodyAccJerk-Mean-Y`                          | Mean of body acceleration jerk in frequency domain along the Y-axis | g (gravity) |
| `FrequencyBodyAccJerk-Mean-Z`                          | Mean of body acceleration jerk in frequency domain along the Z-axis | g (gravity) |
| `FrequencyBodyAccJerk-StandardDeviation-X`             | Standard deviation of body acceleration jerk in frequency domain along the X-axis | g (gravity) |
| `FrequencyBodyAccJerk-StandardDeviation-Y`             | Standard deviation of body acceleration jerk in frequency domain along the Y-axis | g (gravity) |
| `FrequencyBodyAccJerk-StandardDeviation-Z`             | Standard deviation of body acceleration jerk in frequency domain along the Z-axis | g (gravity) |
| `FrequencyBodyGyro-Mean-X`                             | Mean of body gyroscope measurement in frequency domain along the X-axis | rad/s         |
| `FrequencyBodyGyro-Mean-Y`                             | Mean of body gyroscope measurement in frequency domain along the Y-axis | rad/s         |
| `FrequencyBodyGyro-Mean-Z`                             | Mean of body gyroscope measurement in frequency domain along the Z-axis | rad/s         |
| `FrequencyBodyGyro-StandardDeviation-X`                | Standard deviation of body gyroscope measurement in frequency domain along the X-axis | rad/s         |
| `FrequencyBodyGyro-StandardDeviation-Y`                | Standard deviation of body gyroscope measurement in frequency domain along the Y-axis | rad/s         |
| `FrequencyBodyGyro-StandardDeviation-Z`                | Standard deviation of body gyroscope measurement in frequency domain along the Z-axis | rad/s         |
| `FrequencyBodyAccMag-Mean`                             | Mean magnitude of body acceleration in frequency domain | g (gravity)     |
| `FrequencyBodyAccMag-StandardDeviation`                 | Standard deviation of magnitude of body acceleration in frequency domain | g (gravity)   |
| `FrequencyBodyAccJerkMagnitude-Mean`                   | Mean magnitude of body acceleration jerk in frequency domain | g (gravity)   |
| `FrequencyBodyAccJerkMagnitude-StandardDeviation`       | Standard deviation of magnitude of body acceleration jerk in frequency domain | g (gravity)   |
| `FrequencyBodyGyroMagnitude-Mean`                      | Mean magnitude of body gyroscope measurement in frequency domain | rad/s         |
| `FrequencyBodyGyroMagnitude-StandardDeviation`          | Standard deviation of magnitude of body gyroscope measurement in frequency domain | rad/s         |
| `FrequencyBodyGyroJerkMagnitude-Mean`                   | Mean magnitude of gyroscope jerk measurement in frequency domain | rad/s         |
| `FrequencyBodyGyroJerkMagnitude-StandardDeviation`       | Standard deviation of magnitude of gyroscope jerk measurement in frequency domain | rad/s         |

## Data Cleaning and Transformations

1. **Loading Data**:
   - Read training and test datasets along with corresponding inertial signals data files.
   - Dynamically assigned data from each text file to R variables based on filenames.
   
   
```{r}
X_train_path <- "C:\\Users\\bnnen\\Downloads\\getdata_projectfiles_UCI HAR Dataset\\UCI HAR Dataset\\train\\X_train.txt"
y_train_path <- "C:\\Users\\bnnen\\Downloads\\getdata_projectfiles_UCI HAR Dataset\\UCI HAR Dataset\\train\\y_train.txt"
X_test_path <- "C:\\Users\\bnnen\\Downloads\\getdata_projectfiles_UCI HAR Dataset\\UCI HAR Dataset\\test\\X_test.txt"
y_test_path <- "C:\\Users\\bnnen\\Downloads\\getdata_projectfiles_UCI HAR Dataset\\UCI HAR Dataset\\test\\y_test.txt"
activity_labels_path <- "C:\\Users\\bnnen\\Downloads\\getdata_projectfiles_UCI HAR Dataset\\UCI HAR Dataset\\activity_labels.txt"
features_path <-  "C:\\Users\\bnnen\\Downloads\\getdata_projectfiles_UCI HAR Dataset\\UCI HAR Dataset\\features.txt"
subject_train_path <-  "C:\\Users\\bnnen\\Downloads\\getdata_projectfiles_UCI HAR Dataset\\UCI HAR Dataset\\train\\subject_train.txt"
subject_test_path <-  "C:\\Users\\bnnen\\Downloads\\getdata_projectfiles_UCI HAR Dataset\\UCI HAR Dataset\\test\\subject_test.txt"


X_train <- read.table(X_train_path, header = FALSE)
y_train <- read.table(y_train_path, header = FALSE, col.names = "activity_code")
X_test <- read.table(X_test_path, header = FALSE)
y_test <- read.table(y_test_path, header = FALSE, col.names = "activity_code")
features <- read.table(features_path, header = FALSE, col.names = c("index", "feature_name"))
activity_labels <- read.table(activity_labels_path, header = FALSE, col.names = c("activity_code", "activity_name"))
subject_train <- read.table(subject_train_path, header = FALSE, col.names = "subject")
subject_test <- read.table(subject_test_path, header = FALSE, col.names = "subject")


# Function to read data from text files
read_inertial_signals <- function(folder_path) {
  file_list <- list.files(path = folder_path, pattern = "\\.txt$", full.names = TRUE)
  for (file in file_list) {
    file_name <- tools::file_path_sans_ext(basename(file))
    data <- read.table(file, header = FALSE)
    assign(file_name, data, envir = .GlobalEnv)
  }
}

# Read training and test datasets
read_inertial_signals(train_path)
read_inertial_signals(test_path)
```
   

2. **Merging Datasets**:
   - Combined feature data (`X_train`, `X_test`), activity labels (`y_train`, `y_test`), and subject IDs (`subject_train`, `subject_test`) into one comprehensive dataset.
   
   
   
```{r}
# 2. Merge Training and Test Sets
# Combine feature measurements, activity labels, and subject IDs
combined_X <- rbind(X_train, X_test)             
combined_y <- rbind(y_train, y_test)             
combined_subject <- rbind(subject_train, subject_test)

# Set column names for the feature data
colnames(combined_X) <- features$feature_name

# Merge the datasets into one
combined_data <- cbind(combined_subject, combined_y, combined_X)
```
   

3. **Extracting Relevant Features**:
   - Selected only the measurements related to mean and standard deviation by filtering columns containing the keywords "mean()" or "std()".
   - Used the `grep` function to identify and extract these features from the combined dataset.
   
```{r}
# 3. Extract Measurements on Mean and Standard Deviation
mean_std_features <- grep("mean\\(\\)|std\\(\\)", features$feature_name, value = TRUE)
selected_data <- combined_data[, c("subject", "activity_code", mean_std_features)]
```
   

4. **Adding Descriptive Activity Names**:
   - Merged the activity labels with the selected data to replace activity codes with descriptive names.
   
```{r}
# 4. Add Descriptive Activity Names
final_data <- merge(selected_data, activity_labels, by = "activity_code", all.x = TRUE)
final_data <- final_data[, c("subject", "activity_code", "activity_name", mean_std_features)]
```
   

5. **Renaming Variables**:
   - Renamed variables to be more descriptive and user-friendly using the `str_replace_all` function, ensuring clarity (e.g., replacing "mean()" with "Mean" and "std()" with "StandardDeviation").
   
```{r}
# 5. Appropriately Label the Dataset
names(final_data) <- str_replace_all(names(final_data), 
                                     c("mean\\(\\)" = "Mean",
                                       "std\\(\\)" = "StandardDeviation",
                                       "-mean" = "Mean",
                                       "-std" = "StandardDeviation",
                                       "tBodyAcc" = "TimeBodyAcc",
                                       "tGravityAcc" = "TimeGravityAcc",
                                       "tBodyAccJerk" = "TimeBodyAccJerk",
                                       "tBodyGyro" = "TimeBodyGyro",
                                       "tBodyGyroJerk" = "TimeBodyGyroJerk",
                                       "tBodyAccMag" = "TimeBodyAccMagnitude",
                                       "tGravityAccMag" = "TimeGravityAccMagnitude",
                                       "tBodyAccJerkMag" = "TimeBodyAccJerkMagnitude",
                                       "tBodyGyroMag" = "TimeBodyGyroMagnitude",
                                       "tBodyGyroJerkMag" = "TimeBodyGyroJerkMagnitude",
                                       "fBodyAcc" = "FrequencyBodyAcc",
                                       "fBodyAccJerk" = "FrequencyBodyAccJerk",
                                       "fBodyGyro" = "FrequencyBodyGyro",
                                       "fBodyBodyAccJerkMag" = "FrequencyBodyAccJerkMagnitude",
                                       "fBodyAccMag" = "FrequencyBodyAccMagnitude",
                                       "fBodyGyroMag" = "FrequencyBodyGyroMagnitude",
                                       "fBodyBodyGyroMag" = "FrequencyBodyGyroMagnitude",
                                       "fBodyBodyGyroJerkMag" = "FrequencyBodyGyroJerkMagnitude"
                                     ))
```
   

6. **Creating a Tidy Dataset**:
   - Grouped the data by `subject`, `activity_code`, and `activity_name`, then calculated the average of each measurement for each subject and activity using the `summarise` function.
   
```{r}
# 6. Create Tidy Dataset with Averages
tidy_average_data <- final_data %>%
  group_by(subject, activity_code, activity_name) %>%
  summarise(across(starts_with("t") | starts_with("f"), mean, na.rm = TRUE)) %>%
  ungroup()
```
   

7. **Final Dataset**:
   - The final tidy dataset (`tidy_average_data`) contains the average of each variable for each activity and subject.

## Conclusion
The cleaned and transformed dataset provides a comprehensive view of the physical activities performed by subjects, enabling further analysis and insights into human activity recognition. This codebook serves as a guide to understanding the dataset's structure and the transformations applied during the project.
