---
title: "Machine Learning Project2"
author: "Blessing Azubuike"
date: "2024-10-14"
output: html_document
---


## Introduction

The goal of this project is to predict the manner in which participants performed an exercise based on various sensor measurements. The target variable is `classe`, which indicates the type of exercise performed. We will use machine learning techniques to build a predictive model utilizing the training data provided.

## Data Preparation

### Loading Libraries

```{r setup, include=FALSE}
# Load necessary libraries
library(caret)
library(dplyr)
library(randomForest)
library(ggplot2)
```

### Loading Data

```{r}
# Load the training and testing datasets
train_data <- read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv")
test_data <- read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv")

```

### Exploring the Data

```{r}
# Check the structure of the training data
str(train_data)

# Summarize the target variable
table(train_data$classe)

```


### Data Cleaning

```{r}
# Function to remove columns with any NA values
remove_na_columns <- function(data) {
  data[, colSums(is.na(data)) == 0]
}

# Clean both train and test data by removing columns with NA values
train_data_clean <- remove_na_columns(train_data)
test_data_clean <- remove_na_columns(test_data)

# Get the column names of test_data_clean (excluding 'classe' from train_data_clean)
common_columns <- intersect(colnames(test_data_clean), colnames(train_data_clean))

# Keep only common columns in train_data_clean + the 'classe' column
train_data_clean <- train_data_clean[, c(common_columns, "classe")]

# Get the common columns between train_data_clean and test_data_clean, excluding 'classe' from train_data_clean
common_columns <- intersect(colnames(train_data_clean)[colnames(train_data_clean) != "classe"], colnames(test_data_clean))

# Subset the train_data_clean to keep only the common columns and 'classe'
train_data_clean <- train_data_clean[, c(common_columns, "classe")]

# Subset the test_data_clean to keep only the common columns
test_data_clean <- test_data_clean[, common_columns]

# Check if the columns are now aligned
print("Columns in train_data_clean (excluding 'classe'):")
print(colnames(train_data_clean))

print("Columns in test_data_clean:")
print(colnames(test_data_clean))

# Print the dimensions of the cleaned datasets
cat("Dimensions of cleaned training data: ", dim(train_data_clean), "\n")
cat("Dimensions of cleaned testing data: ", dim(test_data_clean), "\n")

```

## Data Cleaning Continued

```{r}
library(dplyr)

# 1. Remove irrelevant columns and those with any NA values
train_data_clean <- train_data %>%
  select(-c(X, user_name, raw_timestamp_part_1, raw_timestamp_part_2, 
            cvtd_timestamp, new_window, num_window)) %>%
  select_if(~ all(!is.na(.)))  # Keep only columns without any NA values

# 2. Convert `classe` to a factor in train_data_clean
train_data_clean$classe <- as.factor(train_data_clean$classe)

# 3. Prepare test_data_clean by removing irrelevant columns and any columns with NA values
test_data_clean <- test_data %>%
  select(-c(X, user_name, raw_timestamp_part_1, raw_timestamp_part_2, 
            cvtd_timestamp, new_window, num_window)) %>%
  select_if(~ all(!is.na(.)))  # Keep only columns without any NA values

# 4. Ensure both datasets have the same columns (except `classe` in train_data_clean)
common_columns <- intersect(colnames(train_data_clean)[colnames(train_data_clean) != "classe"], colnames(test_data_clean))

# Subset the train_data_clean to keep only the common columns and 'classe'
train_data_clean <- train_data_clean[, c(common_columns, "classe")]

# Subset the test_data_clean to keep only the common columns
test_data_clean <- test_data_clean[, common_columns]

# Check if the columns are now aligned
print("Columns in train_data_clean (excluding 'classe'):")
print(colnames(train_data_clean))

print("Columns in test_data_clean:")
print(colnames(test_data_clean))

# Check the levels and number of observations for each class in the training data
print("Levels of 'classe':")
print(levels(train_data_clean$classe))
print("Distribution of 'classe' in training data:")
print(table(train_data_clean$classe))
```

### Model Building

#### Train-Test Split

```{r}
# Set seed for reproducibility
set.seed(123)

# Create training and validation sets
train_index <- createDataPartition(train_data_clean$classe, p = 0.75, list = FALSE)
training_set <- train_data_clean[train_index, ]
validation_set <- train_data_clean[-train_index, ]

# Check for factor levels in the training set
print("Levels of 'classe' in training set:")
print(levels(training_set$classe))
print("Distribution of 'classe' in training set:")
print(table(training_set$classe))

```

### Model Training and cross validation

```{r}
# Ensure we have enough levels for the response variable before training the model
if (length(levels(training_set$classe)) > 1) {
  # Train the random forest model
  rf_model <- randomForest(classe ~ ., data = training_set, importance = TRUE, ntree = 100)
  
  # Print the random forest model summary
  print(rf_model)
  
  # Train the random forest model with cross-validation
  train_control <- trainControl(method = "cv", number = 10)
  
  cv_model <- train(classe ~ ., data = training_set, method = "rf", trControl = train_control)
  
  # Print the model results
  print(cv_model)
  
} else {
  stop("Not enough levels in 'classe' for model training.")
}

```


### Model Evaluation

#### Accuracy of the Model

```{r}
# Evaluate the model on the validation set
predictions <- predict(rf_model, validation_set)
confusionMatrix(predictions, validation_set$classe)

```


#### Expected Out-of-Sample Error

Based on the confusion matrix, we can assess the model's accuracy and derive the expected out-of-sample error

### Predictions on Test Data

```{r}
# Make predictions on the test dataset
test_predictions <- predict(rf_model, test_data_clean)

# Display predictions
print(test_predictions)


```


### Conclusion

In this report, we have built a predictive model using the random forest algorithm to predict the manner in which exercises were performed. We used cross-validation to evaluate the model's performance and derived an expected out-of-sample error. The model performed well on the validation set with an accuracy of 0.9963, and we also made predictions for the provided test cases.






















